Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-21 demystify.tex
updates

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{soul}
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\newcommand{\myepigraphsimpsons}[1]{
\epigraph{\vspace{-0.2in} \emph{#1}}{\textsc{The Simpsons}}
}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float

\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{xcolor}
%\usepackage{anyfontsize}
%\usepackage{sectsty}

@@

\newtcolorbox{myemphasisbox}{
  enhanced,
  boxrule=1pt, frame hidden,
  colback=gray!20!white,
  sharp corners
}

@@

\newcommand{\chapterinfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{chapterinfo}{\textcolor{black}{\emph{#1}}}%
}
\newcommand{\sectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{sectioninfo}{\textcolor{black}{\emph{#1}}}%
}
\newcommand{\subsectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{subsectioninfo}{\textcolor{black}{\texttt #1}}%
}

@@

\usepackage{hyperref}
\hypersetup{
    hidelinks,
    linktoc=all,
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
  pdfpagemode=UseOutlines,
}

\usepackage[margin=1.5in]{geometry}

@@

\newpage
\renewcommand{\contentsname}{Contents and Summary}
\tableofcontents

@@

\part{Introduction To CS PhD Admission}\label{part:intro}

@@

\chapterinfo{Yes, definitely.  CS PhD study in the US is often fully funded and admission into good universities in the US is not any harder than in other countries.}

\myepigraphsimpsons{``Don't make fun of graduate students. They just made a terrible life choice.''}

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.
Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This document aims to address these concerns.

@@

\paragraph{Discrimination is not tolerated} Third, not only is discrimination illegal, diversity and inclusion are highly valued in US universities. Many universities take great pride in their diverse student body and faculty.

In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also \emph{very} familiar with and are highly encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

The guideline can be useful for all students, but it is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research opportunities).

Students from top schools with strong research programs and experience might already know most of the information given here.
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply to CS PhD programs in the US.

Another goal of this document is to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

While this document may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions. It is certainly useful for students aiming for Mason!

@@

No, while other countries often require an MS for PhD student in CS (~\autoref{sec:non-us-differences}), it is common in he US to apply for a PhD program directly after a 4-year undergrad program (e.g., after getting a B.S degree).  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}

\sectioninfo{Among several differences, CS PhD in the US does not require an MS degree but has a longer PhD study time.}

\begin{table}
\caption{Comparison of the CS PhD program in the US and other countries}\label{tab:us-vs-other}
\center
\small
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (\textbf{adcom}) committee and each application is typically reviewed by three faculty members.}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application. They consider various factors, e.g., research experience, LoRs, SOP}

Your applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. Adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. The adcom \emph{chair} will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  They will consider various factors, e.g., research experience, GPA, LoRs, SOP, and test scores. More details about these factors are discussed in \autoref{part:appendices}.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak LoRs.
\end{commentbox}

@@

Note that while the assigned reviewers are the main ones deciding your application, other faculty in the department can also have access to your application and provide inputs and opinions on your profile. Thus, it helps to contact faculty (\autoref{sec:contact}) and mention faculty you're interested in in your SOP (\autoref{chap:sop}).

@@

\section{Why CS depts do not waive the application fee?}\label{sec:fee-waive}

@@

\part{Application Materials}\label{part:application}

@@

\myepigraphsimpsons{``Son, if you really want something in this life, you have to work for it. Now quiet! They’re about to announce the lottery numbers.''}

@@

\chapter{Letters of Recommendation (LoR)}\label{chap:lor}
\chapterinfo{LORs are very important, but only if they are \textbf{personalized} and \textbf{research focused}.}

\myepigraphsimpsons{``To whom it may concern… D’oh!''}

@@

Most PhD programs will require at least \emph{two} letter of recommendations (LoRs). 
The goal of the LoRs is to provide the adcom with an assessment of your research ability from someone who knows you.  The LoRs should focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher to adcom.

\section{LoR writers}\label{sec:lor-writers} \sectioninfo{LoR writers should be someone who (i) can talk in depth about your research experience and potential and (ii) have the credibility to evaluate your research ability.}

@@

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also \emph{have the credibility} to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD, so they know what a PhD is about and so can evaluate your research ability. LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).

@@

%It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This also applies to administrative, e.g., chairs, directors, and deans (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

  \subsection{LoRs from Dept Chair, Dean, or Boss and Supervisor at Work}\label{sec:admin-letters}

  Many students try to get LoRs from high-ranking administrators in their universities such as department chair/head, dean, or director. The students never worked with these people (they might take a class or so with these profs), but mistakenly believe that these LoRs are valuable due to the writer's high position in the university.
  However, as emphasized, a generic LoR has little value because the writer does not know you well and can talk in depth about your research ability.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}

Many profs. and letter writers ask students to write their own letters, which is a common practice in some countries. Unfortunately, such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in it is you who write about yourself (and they just sign the letter).

Self-written letters are easy to spot as an experienced professor would write very differently compared to an undergraduate student.
For example, they can provide convincing and concrete examples based on their experience and compare you to their own students, and of course the writing style of a professor is very different from an undergraduate student.  Worse yet, if we suspect that the student wrote the letter, it is a \red{red flag} as we will question both the student's integrity and the letter writer's credibility.

@@

\section{Asking for LORs}\label{sec:asking-lor}
\sectioninfo{Ask for LoRs at least a month before the deadline.  Waive your right to see the letter. Help your writers by providing details about your research experience and potential.}

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  In the US, students usually explicitly ask if the writer would be willing to provide a strong letter, and the writer are also very direct in their response.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

Below are some tips to approach your LoR writers:
\begin{itemize}
\item \textbf{Ask in advance} You should ask for LoRs \emph{at least a month} before the deadline. People are busy and writing a strong LoR takes time (\autoref{sec:lor-writing}), so give them enough time to write a strong letter for you.

\item \textbf{Waive your right (\autoref{sec:waive-right})} You should always waive your right to see the letter. This shows that you trust your writers and that you are not trying to twist their words.

\item \textbf{Help your writers (\autoref{sec:help-your-LOR-writers})}: You should tell your writers the programs you are applying to, their deadlines, etc.  You can also share your SOP with them and other details about your research experience and potential.

\item \textbf{Ask for feedback} If the writer is very close to you, you can ask them for feedback on your SOP (\autoref{chap:sop}) and other application materials. If the writer is a professor, they might have served in adcom committees, seen many SOPs, and can provide valuable feedback.

\item \textbf{Follow up and Stay in touch} Follow up with your writers to make sure they have submitted the letters on time. Note that their letters might have a different due date than your application (\autoref{sec:remind-writers}).

After your writers have submitted all of their letters, \emph{thank them} (\autoref{sec:thank-writers}).  Finally, let them know the outcome of your applications and stay in touch with them.  This will help you build a relationship with them and you might need them to write for you again in the future.
\end{itemize}

@@

When you ask someone to write a letter for you, you should \emph{always} waive your right.

@@

\subsection{Helping your LOR Writer}\label{sec:help-your-LOR-writers}

 As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you can provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\subsection{Reminding Your Writers}\label{sec:remind-writers}

After entering your writers' information in the application system, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also tell them when you submit your application and remind them to submit their letters on time if they haven't done so.

@@

\subsection{Thank Your Writers}\label{sec:thank-writers}
After your application is submitted and your writers have submitted their letters (i.e., the wait begins), you should send a quick thank you note, which serves both as an acknowledgement that you know they have submitted the letters and as an appreciate for their help.

You should also let them know the outcome of your application, regardless of whether you are admitted or not.  In addition to being a common courtesy, this can also help maintain a good relationship with your writers, which can be useful in the future (e.g., if you need another letter for another round of applications or for a job reference).

@@

I will also let the student know if I cannot write a strong letter for them (e.g., I don't know them that well), and suggest they find someone else.  If they insist, then I will write for them. While I try to say something positive, e.g., the student is hardworking, or that I heard they are working on a research project, etc, the letter will still be short and weak (\autoref{sec:generic-letters}). Usually it takes me about \emph{5--10} minutes to write such a letter.

Strong letters take a lot longer as it will be personalized.  While I have a general template for LoR, it still can take me \emph{an hour or more} to write it.  I often ask the students to provide information (\autoref{sec:help-your-LOR-writers}) and what they think I should highlight in the letter. They can also provide me their SOP (if they already have written one) so I can complement what they say with my own perspective.  I often do not share what I write with the students, just to keep it more genuine and honest.

My letter always has the university letterhead, my signature, and is signed and dated.  While using a letterhead or signing the letter is not required (I never paid much attention to these when reading LoRs), it makes the letter more official and professional.  
This website has the template of the letter with letterhead and signature: \url{https://www.overleaf.com/read/xzyrxkdjfxsp#2a1f9e}.

@@

\chapter{Research Experience}\label{chap:research-experience}
\chapterinfo{Publications are not \textbf{required} but can \textbf{greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research.}

@@

\section{Publications}\label{sec:publications}

The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that you have been successfully involved in research.

@@

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.  
So do talk about them in your SOP (\autoref{chap:sop}) and have your LoR writers mention them in their letters (\autoref{chap:lor}).

\chapter{Statement of Purpose (SOP)}\label{chap:sop}
\chapterinfo{SOP is important. Write it in such a way that makes you \textbf{stand out}.}

@@

In your SOP, focus on research potential (\autoref{chap:research-experience}) and convince us through your experience, e.g., published papers (\autoref{sec:publications}). Back up your claims with \textbf{concrete evidence}. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a research project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art.

@@

Finally, have your SOP reviewed by your LoR writers (\autoref{sec:help-your-LOR-writers}) and professors, especially those who have served in adcom, or even postdocs or PhD students as they have been through this process.

@@

\begin{commentbox}[Vu:]
One common pitfall is to write about how you are from a poor family, a small town, or a developing country.  These are not related to DEI and show that you do not understand the topic.   I have seen people consider themselves as minority because they are Asian or international students studying in the US (have you seen how many Asian or international students are in CS programs?).\\

Moreover, do not make up stories or exaggerate your experiences.  Just show your awareness and understanding of the topic and what you did or how you plan to contribute to it (e.g., mentoring students, helping others). One of the best DEI statements I have read starts the acknowledgement that the applicant is not a minority, but a white male in a STEM field (i.e., they are the majority), and that they do not have much experience in DEI, but show that they are aware of the challenges faced by minority groups and how they can contribute to it.
\end{commentbox}

\chapter{Your School and Grades}\label{chap:your-school}
\chapterinfo{High grades probably won't help much, but bad ones likely will raise concerns.}

@@

\chapter{Standard Tests (GRE and IELTS)}\label{chap:standard-tests}
\chapterinfo{Standard tests are \textbf{not} important. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a good sign; but no interview does not mean rejection}

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{pdfpages}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}

\usepackage{tikz}
\usetikzlibrary{calc}
\usepackage{xcolor}

\usepackage{anyfontsize}
\usepackage{sectsty}

@@

\newtcolorbox{myemphasisbox}{
  enhanced,
  boxrule=1pt, frame hidden,
  %borderline = {1pt}{0pt}{blue!75!black},
  colback=gray!20!white,
  sharp corners
}

@@

\def\chapterinfo#1{%
  \addcontentsline{toc}{chapterinfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}
\def\sectioninfo#1{%
  \addcontentsline{toc}{sectioninfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}
\def\subsectioninfo#1{%
    \addcontentsline{toc}{subsectioninfo}{%
    \noexpand\numberline{}\color{black}{#1}}%
}

@@

@@

\newpage
\tableofcontents

% \chapter{Summary}\label{sec:summary}

% This section summarizes the main points of this guideline. It also gives you an overview to decide which specific topics you want to explore more thoroughly.

  %\item Should you apply to a PhD CS program in the US?

          %\item \emph{Contacting a prof. is recommended}, but do it \emph{properly} to get a reply (\autoref{sec:contact}).

          %\item  (\autoref{sec:interview}).

    % \item Miscs and FAQS
    %     \begin{itemize}

    %       \item (\autoref{sec:read-all}).

    %       \item
    %       \item
    %       \item (\autoref{sec:msrequirement}), and CS PhD takes abou (\autoref{sec:time}). Pu(\autoref{sec:non-us-differences}).
    %       \item
    %       \item
    %       \item
    %       %\item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
    %     \end{itemize}

    % \item Appendices
    % \begin{itemize}
    %   \item Specific advice for domestic students (\autoref{sec:domestic-students}).
    %   \item How to find research opportunities (\autoref{sec:research-opportunities}).
    %   \item Cultural differences between US and other countries (\autoref{sec:cultural}).
    % \end{itemize}

@@

\part{Introduction To CS PhD Admission in the US}\label{part:intro}

@@

\chapterinfo{\emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities in the US is not any harder than in other countries.}

\epigraph{\vspace{-0.2in} Don't make fun of graduate students. They just made a terrible life choice.}{\textsc{The Simpsons}}

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} based on gender identity, sexual orientation, or cultural background.

Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This document aims to address these concerns.

@@

\paragraph{Discrimination is not tolerated} Third, higher education in the US take a very strong stance against discrimination, especially at the university level.
Many resources are available to encourage and support students from minority groups to pursue research, e.g., specific scholarships and fellowships. Faculty in CS are also \emph{very} familiar with and are highly encouraged to support such students in their group. There are also incentives, such as dedicated funding, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

While this guideline can be useful for all students, it would most benefit \textbf{international students} from \textbf{smaller countries and less well-known universities} (or domestic students from US universities with no PhD programs or limited research opportunities).

Students from top schools with strong research programs and experience might already know most of the information given here. They might also receive guidance from their professors or grad students who have gone through the process. Students from smaller schools or countries, however, might not have access to such resources and might not know where to start. 
My goal is thus to level the playing field by providing info that is not readily available to less privileged students. 
I hope to encourage more students with such backgrounds to apply and succeed in CS PhD programs in the US.

This document also aim to explain the \emph{``why''} behind the admission process, e.g., why LoRs are important, why you should not draft your own LoR, why you should contact professors, etc. There are numerous \href{https://github.com/nguyenthanhvuh/phd-cs-us}{resources online} from students  that tell you \emph{``what''} to do, but few explain the \emph{``why''}. Understanding the reason and mindset of the admission committee and profs. can help you prepare better.

While this document may not fully meet the needs of students aiming for top-10 schools (\autoref{sec:selecting-ranking-schools} and \autoref{sec:ranking}), which typically have more stringent requirements (e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU), it should be valuable for those targeting other reputable R1 institutions, which are frankly the majority of good schools in the US.

@@

No, not in the US (other countries often do though~\autoref{sec:non-us-differences}).  In the US, it is common to apply for a PhD program directly after your undergrad.  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\section{PhD in the US vs. Other Countries}\label{sec:non-us-differences}
\sectioninfo{Among several differences, CS PhD in the US does not require an MS degree but has a longer PhD study time.}

\begin{table}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\center
\small
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\sectioninfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

\epigraph{How is education supposed to make me feel smarter? Besides, every time I learn something new, it pushes some of the old stuff out of my brain. Remember when I took that home wine making course, and I forgot how to drive?}{\textsc{The Simpsons}}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}

Your applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

Each application is assigned to about three adcom members, who will evaluate your profile and reach a consensus.  While the assigned reviewers are the main ones deciding your application, other faculty in the department can also have access to your application and provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. The chair of the committee will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

\begin{commentbox}{Vu:}
At Mason, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision (\autoref{sec:ievaluate}) with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.
\end{commentbox}

@@

@@

\subsection{Why CS depts do not waive the application fee?}\label{sec:fee-waive}

@@

\part{Your Application Materials}\label{part:application}

@@

\epigraph{Son, if you really want something in this life, you have to work for it. Now quiet! They’re about to announce the lottery numbers.}{\textsc{The Simpsons}}

@@

\chapter{Letters of Recommendation (LOR)}\label{chap:lor}
\chapterinfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

Most PhD programs will require at least \textbf{two LORs}. 
The goal of the LoR is to provide the adcom with an assessment of your research ability and potential. 
The letter should 
focus on your research background, achievements, and potentials, all of which paint a compelling picture of you as a researcher.

\section{LoR writers}\label{sec:lor-writers}

@@

A LoR writer should be someone who \emph{knows you well}, so that they can provide a strong-supporting and personalized letter for you.  LoR writers should also \emph{have the credibility} to convince adcom members on your research ability and potential. Such a person is typically someone who has a PhD (so they know what a PhD is about) and is active in research (so they can evaluate your research ability). LoR writers are often your research advisers, professors, or supervisors who have worked with you on research projects.

\subsection{LoR from Famous People}  Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong LOR from such a researcher can outweigh the lack of publications (or low GPA or unknown school). The reason is simple: adcom members trust people that they know and respect. If a well-known researcher enthusiastically recommends you, then the adcom members will likely trust that recommendation.

@@

It is also worth noting that a LoR from a ``full'' senior professor who has not been active in research might not be as strong as a LoR from a junior faculty who has been actively mentoring you in research.  This also applies to administrative, e.g., chairs, directors, and deans (\autoref{sec:admin-letters}.) The fame of the writer is less important than how well they know you and can write about your research ability.

@@

  \subsection{LoRs from Admin People}\label{sec:admin-letters}

  Many international students try to get LoRs from high-ranking administrators in their universities such as department chair/head, dean, or director. The students never worked with these people (they might take a class or so with these profs), but mistakenly believe that these LoRs are valuable due to the writer's high position in the university.
  However, as emphasized, a generic LoR has little value because the writer does not know you well and can talk in depth about your research ability.

@@

\subsection{Self-written Letters are Bad}\label{sec:self-letters}
Many students write letters themselves, typically due to the requests of their profs. or letter writers. Such letters have \emph{little value} and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the ref. writer talking about you, in it is you who write about yourself (and they just sign the letter).

Adcom members are just too experienced (having reviews hundreds of those letters) to spot and suspect such letters.  There is a very difference between a letter written by an experienced professor and a student.  For example, they can provide convincing and concrete examples based on their experience and compare you to their own students, and of course the writing style of a professor is very different from an undergraduate student.  Worse yet, if we suspect that the student wrote the letter, it is a \red{red flag} as we will question both the student's integrity and the letter writer's credibility.

@@

\section{Asking for LORs}\label{sec:asking-lor}

As mentioned in~\autoref{sec:lor-writers}, LoR writers should be someone who knows you well and has the credibility to evaluate your research ability.  You can explicitly ask if they are willing to write a strong letter for you.  If they are not willing (you should be thankful that they are honest with you), then you should ask someone else.

% Below are some tips to approach your LoR writers:
% \begin{itemize}
% \item \textbf{Ask in advance}: You should ask for LoRs \emph{at least a month} before the deadline. People are busy with their own work and likely have many other students asking for LoRs. Also, a strong LoR takes time to write, so you should give them enough time to write a strong letter for you.

% \item \textbf{Help your writers}: You should provide your writers with details about the programs you are applying to, the deadlines, and any other information they need to know.  You should also provide them with details about your research experience and potential so that they can write a strong letter for you.
% \item \textbf{Follow up}: After you ask for LoRs, you should follow up with your writers to make sure they have submitted the letters on time.  You should also thank them for writing the letters for you.
% \item \textbf{Waive your right}: When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should always waive your right.  This shows that you trust your writers and that you are not trying to twist their words.
% \item \textbf{Remind your writers}: After you enter your writers' information in the application system, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.
% \item \textbf{Help your writers}: To help your writers write a strong letter for you, you should provide them with details about your research experience and potential.  You should also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
% \item \textbf{Thank your writers}: After your writers have submitted their letters, you should thank them for writing the letters for you.  You should also let them know the outcome of your applications.
% \item \textbf{Ask for feedback}: After you have been admitted or rejected, you should ask your writers for feedback on your application.  This will help you improve your application for future applications.
% \item \textbf{Stay in touch}: After you have been admitted or rejected, you should stay in touch with your writers.  This will help you build a relationship with them and will help you in the future.
% \end{itemize}

@@

When you ask someone to write a letter for you, you will be asked if you want to waive your right to see the letter.  You should \emph{always} waive your right.

@@

\subsection{Help your LOR Writer}\label{sec:help-your-LOR-writers}

 As mentioned above in \autoref{sec:self-letters} and \autoref{sec:generic-letters}, do not write your own letter and generic letters do not give much value. Thus, to help your writer to write a strong, customized LoR, you should provide them details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention.  If the GPA in your program is highly competitive (\autoref{sec:gpa}) and they know that, remind them to talk about it in the LOR. 
You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.

@@

\subsection{Remind Your Writers}\label{sec:remind-writers}

After entering your writers' information in the application system, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.

@@

\paragraph{Thank Your Writers} After your application is submitted and your writers have submitted their letters (i.e., the wait begins), you should send a quick thank you note, which serves both as an acknowledgement that you know they have submitted the letters and as an appreciate for their help.  You should also let them know the outcome of your application, regardless of whether you are admitted or not.  In addition to being a common courtesy, this can also help maintain a good relationship with your writers, which can be useful in the future (e.g., if you need another letter for another round of applications or for a job reference).

@@

I will also let the student know if I cannot write a strong letter for them (e.g., I don't know them that well), and suggest they find someone else.  If they insist, then I will write for them. While I try to say something positive, e.g., the student is hardworking, or that I heard they are working on a research project, etc, the letter will still be short and weak (\autoref{sec:generic-letters}). Usually it takes me about 5--10 minutes to write such a letter.

For a student that I know well, I will write them a strong letter, which takes a lot longer as it will be personalized.  While I have a general template for LoR, it still can take me an hour or more to write it.  I often ask the students to provide information (\autoref{sec:help-your-LOR-writers}) and what they think I should highlight in the letter. They can also provide me their SOP (if they already have written one) so I can complement what they say with my own perspective.  I often do not share what I write with the students, just to keep it more genuine and honest.

@@

\chapter{Research Experience}\label{chap:research-experience}

\sectioninfo{\emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required!}

@@

\section{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that you have been successfully involved in research.

@@

\section{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

\chapter{Statement of Purpose (SOP)}\label{chap:sop}
\chapterinfo{\emph{SOP is important}. Write it in such a way that makes you \emph{stand out}.}

@@

In your SOP, focus on research potential and convince us through your experience, e.g., published papers. Back up your claims with \textbf{concrete evidence}. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on a project, then show some results, e.g., paper submitted (or even rejected), achieved certain performance improvement over the state of the art.

@@

Finally, your SOP reviewed by your LoR writers and professors, especially those who have served in adcom, or even postdocs or PhD students as they have been through this process.

@@

\begin{commentbox}[Vu:]
One common pitfall is to write about how you are from a poor family, a small town, or a developing country.  These are not related to DEI and raise concerns about you not understanding or even bother to do some basic research about the topic.   I have seen people consider themselves as minority because they are Asian or international students studying in the US (have you seen how many Asian or international students are in CS programs in the US?).\\

Moreover, do not make up stories or exaggerate your experiences.  Just show your awareness and understanding of the topic and what you did or how you plan to contribute to it (e.g., mentoring students, helping others). One of the best DEI statements I have read starts the acknowledgement that the applicant is not a minority, but a white male in a STEM field (i.e., that make them the majority), and that they do not have much experience in DEI, but show that they are aware of the challenges faced by minority groups and how they can contribute to it.
\end{commentbox}

\chapter{Your School and Grades}\label{chap:your-school}
\chapterinfo{\emph{High grades probably won't help much} (unless you're from a very top and well-known school), but bad ones likely will raise concerns.}

@@

\chapter{Standard Tests (GRE and IELTS)}\label{chap:standard-tests}
\chapterinfo{\emph{Standard tests are not important}. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection}
